Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-07-01 Manuscript/manuscript.Rmd
edits

After determining the optimal set of data level characteristics, we then aimed to compare the performances of annotation-specifc models, within each cell line and across resolution. In doing so, we inspected the proportion of predictive importance that was explained by each of the 3 groups of annotations (chromatin states, histone modifications, and transcription factor binding sites). We found that for cell line specific random forest models implemented on distance-type predictors, using SMOTE re-sampling, the majority of the proportional predictive importance was obtained by the functional genomic annotations classified as transcription factor binding sites (TFBS) (Figure 6). This was consistent across resolutions, suggesting that TFBS were the primary drivers of TAD boundary formation in both the GM12878 and K562 cell lines.

We further assessed the predictive capabilities of the specific sets of annotations by comparing the performances of annotation-specific random forest models implemented on the optimal set of characteristics. We found that cell line specific models built simply on TFBS out performed other cell line and annotation-specific models at all resolutions (Figure 7). Additionally, when compared to models built on all 3 sets of functional genomic annotations, the TFBS-specific models performed similarly at the 10 kb and 25 kb resolutions. Interestingly though, at the 50 kb and 100 kb resolutions, the TFBS-specific models out performed models built on the combined sets of annotations for both cell lines. This indicated that, at lower resolutions, the inclusion of other types of annotations contributed to additional noise rather than signal.

@@

We then sought to determine which of the transcription factor binding sites were most influential in predicting TAD boundaries. We found that the cell line specific binding sites associated with SMC3, CTCF, RAD21, ZNF143, and YY1sc281 were given consistently high predictive importance, across all resolutions (Figure 7; Supplementary Figure 6). The SMC3, CTCF, and RAD21 binding sites are known factors associated with the cohesin protein complex, which is hypothesized to be what creates loops in the DNA  [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likewise, ZNF143 and YY1 are known structural regulators of enhancer-promoter interactions that also contribute to looping of the linear genome [@weintraub2017yy1; @gonzalez2017znf143]. With the identification of these factors known to be enriched at TAD boundaries, we are able to confirm the validity of our model building process. Additionally, we were also able to identify some lesser known cell line specific transcription factors as predictive of TAD boundaries including the MYC-associated zing-finger protein (MAZ), jun-D proto-oncogene (JUND), and ELF1. These transcription factor binding sites tended to be given higher predictive importance at lower resolutions. In fact, at lower resolutions, the predictive importance was found to be much more evenly dispersed across the top transcription factor binding sites, than for higher resolutions. This is again suggestive of the fact that the spatial relationship between functional annotations, specifically transcription factor binding sites, and TAD boundaries changes at different Hi-C data resolutions.

>>>>

After determining the optimal set of data level characteristics, we then aimed to compare the performances of annotation-specifc models, within each cell line and across resolution. In doing so, we inspected the proportion of predictive importance that was explained by each of the 3 groups of annotations (chromatin states, histone modifications, and transcription factor binding sites). We found that for cell line specific random forest models implemented on distance-type predictors, using SMOTE re-sampling, the majority of the proportional predictive importance was obtained by the functional genomic annotations classified as transcription factor binding sites (TFBS) (Figure 6 A). This was consistent across resolutions, suggesting that TFBS were the primary drivers of TAD boundary formation in both the GM12878 and K562 cell lines.

We further assessed the predictive capabilities of the specific sets of annotations by comparing the performances of annotation-specific random forest models implemented on the optimal set of characteristics. We found that cell line specific models built simply on TFBS out performed other cell line and annotation-specific models at all resolutions (Figure 6 B). Additionally, when compared to models built on all 3 sets of functional genomic annotations, the TFBS-specific models performed similarly at the 10 kb and 25 kb resolutions. Interestingly though, at the 50 kb and 100 kb resolutions, the TFBS-specific models out performed models built on the combined sets of annotations for both cell lines. This indicated that, at lower resolutions, the inclusion of other types of annotations contributed to additional noise rather than signal. These results imply that restricting the feature space of downstream predictive models to transcription factor binding sites is less computationally intesive and does not necessarily sacrifice performance.

@@

We then sought to determine which of the cell line specific transcription factor binding sites were most influential in predicting TAD boundaries.

We found that the cell line specific binding sites associated with SMC3, CTCF, RAD21, ZNF143, and YY1sc281 were given consistently high predictive importance, across all resolutions (Figure 6 C). The SMC3, CTCF, and RAD21 binding sites are known factors associated with the cohesin protein complex, which is hypothesized to be what creates loops in the DNA  [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likewise, ZNF143 and YY1 are known structural regulators of enhancer-promoter interactions that also contribute to looping of the linear genome [@weintraub2017yy1; @gonzalez2017znf143]. With the identification of these factors known to be enriched at TAD boundaries, we are able to confirm the validity of our model building process. Additionally, we were also able to identify some lesser known cell line specific transcription factors as predictive of TAD boundaries including the MYC-associated zing-finger protein (MAZ), jun-D proto-oncogene (JUND), and ELF1. These transcription factor binding sites tended to be given higher predictive importance at lower resolutions. In fact, at lower resolutions, the predictive importance was found to be much more evenly dispersed across the top transcription factor binding sites, than for higher resolutions. This is again suggestive of the fact that the spatial relationship between functional annotations, specifically transcription factor binding sites, and TAD boundaries changes at different Hi-C data resolutions.
